Simple repression draft

### 2020-09-10 doc/resubmission/section_02_means.tex
implement Rob's comments

\fig{fig1:means_cartoons} depicts a smorgasbord of mathematicized cartoons for
simple repression using both thermodynamic and kinetic models that have appeared
in previous literature. For each cartoon, we calculate the fold-change in mean
gene expression as predicted by that model, deferring some algebraic details to
Appendix~\ref{sec:non_bursty}. What we will find is that all cartoons collapse
to a single master curve, shown in \fig{fig1:means_cartoons}(D), which contains
just two parameters. We label the parameters $\Delta F_R$, an effective free
energy parametrizing the repressor-DNA interaction, and $\rho$, which subsumes
all details of transcription in the absence of repressors. We will offer some
intuition for why this master curve exists and discuss why at the level of the
mean expression, we are unable to discriminate ``right'' from ``wrong'' cartoons
given only measurements of fold-changes in expression.

@@

Again, we spare all details of the derivation to Appendix~\ref{sec:non_bursty}.
Here we just highlight the general findings for all five kinetic models. As
already shown in Figure~\ref{fig1:means_cartoons}(C) and (D), all the kinetic
models explored can be collapsed into the master curve. Given that the
repressor-bound state only connects to the rest of the promoter dynamics via its
binding and unbinding rates, $k_R^+$ and $k_R^-$ respectively, all models can
effectively be separated into two categories: a single repressor-bound state,
and all other promoter states with different levels of coarse graining. This
structure then guarantees that, at steady-state, detailed balance between these
two groups is satisfied. What this implies is that the steady-state distribution
of each of the non-repressor states has the same functional form with or without
the repressor, allowing us to write the fold-change as a product of the ratio of
the binding and unbinding rates of the promoter, and the promoter details. This
is 
\mrm{This is based in one of the comments from reviewer \# 2. It is empirically
true for all models we explored, but I don't know how to prove that it is a 
general statement for any unimaginable complex model.}
\begin{align}
\text{fold-change} &= \left( 1 + \frac{k_R^+}{k_R^-} \rho \right)^{-1},\\
&= (1 + \exp(-\Delta F_R + \log(\rho) ))^{-1},
\end{align}
where $\Delta F_R \equiv -\log(k_R^+/k_R^-)$, and the functional forms of $\rho$
for each model change as shown in Figure~\ref{fig1:means_cartoons}(C) (See
Appendix~\ref{sec:non_bursty} for further details).

The key outcome of our analysis of the models in
Figure~\ref{fig1:means_cartoons} is the existence of a master curve shown
in Figure~\ref{fig1:means_cartoons}(D) to which the fold-change predictions
of all the models collapse. This master curve is parametrized by only two
effective parameters: $\Delta F_R$, which characterizes the number of repressors
and their binding strength to the DNA, and $\rho$, which characterizes all other
features of the promoter architecture. The key assumption underpinning this
result is that no transcription occurs when a repressor is bound to its
operator. Given this outcome, a mean-based metric such as the fold-change that
can be readily measured experimentally is insufficient to discern between these
different levels of coarse-graining. In the next section we extend the analysis
of the models to higher moments of the mRNA distribution as we continue to look
for the ``right'' level of coarse graining.

>>>>

\fig{fig1:means_cartoons} depicts a smorgasbord of mathematicized cartoons for
simple repression using both thermodynamic and kinetic models that have appeared
in previous literature. For each cartoon, we calculate the fold-change in mean
gene expression as predicted by that model, deferring some algebraic details to
Appendix~\ref{sec:non_bursty}. What we will find is that all cartoons collapse
to a single master curve as shown in \fig{fig1:means_cartoons}(D), which
contains just two parameters. We label the parameters $\Delta F_R$, an effective
free energy parametrizing the repressor-DNA interaction, and $\rho$, which
subsumes all details of transcription in the absence of repressors. We will
offer some intuition for why this master curve exists and discuss why at the
level of the mean expression, we are unable to discriminate ``right'' from
``wrong'' cartoons given only measurements of fold-changes in expression.

@@

Again, we consign all details of the derivation to Appendix~\ref{sec:non_bursty}.
Here we just highlight the general findings for all five kinetic models. As
already shown in Figure~\ref{fig1:means_cartoons}(C) and (D), all the kinetic
models explored can be collapsed onto the master curve. Given that the
repressor-bound state only connects to the rest of the promoter dynamics via its
binding and unbinding rates, $k_R^+$ and $k_R^-$ respectively, all models can
effectively be separated into two categories: a single repressor-bound state,
and all other promoter states with different levels of coarse graining. This
structure then guarantees that, at steady-state, detailed balance between these
two groups is satisfied. What this implies is that the steady-state distribution
of each of the non-repressor states has the same functional form with or without
the repressor, allowing us to write the fold-change as a product of the ratio of
the binding and unbinding rates of the promoter, and the promoter details. This
is 
\mrm{This is based in one of the comments from reviewer \# 2. It is empirically
true for all models we explored, but I don't know how to prove that it is a 
general statement for any unimaginable complex model.}
\begin{align}
\text{fold-change} &= \left( 1 + \frac{k_R^+}{k_R^-} \rho \right)^{-1},\\
&= (1 + \exp(-\Delta F_R + \log(\rho) ))^{-1},
\end{align}
where $\Delta F_R \equiv -\log(k_R^+/k_R^-)$, and the functional forms of $\rho$
for each model change as shown in Figure~\ref{fig1:means_cartoons}(C) (See
Appendix~\ref{sec:non_bursty} for further details).

The key outcome of our analysis of the models in
Figure~\ref{fig1:means_cartoons} is the existence of a master curve shown
in Figure~\ref{fig1:means_cartoons}(D) to which the fold-change predictions
of all the models collapse. This master curve is parametrized by only two
effective parameters: $\Delta F_R$, which characterizes the number of repressors
and their binding strength to the DNA, and $\rho$, which characterizes all other
features of the promoter architecture. The key assumption underpinning this
result is that no transcription occurs when a repressor is bound to its
operator. Given this outcome, i.e., the degeneracy of the different models at
the level of fold-change, a mean-based metric such as the fold-change that can
be readily measured experimentally is insufficient to discern between these
different levels of coarse-graining. In the next section we extend the analysis
of the models to higher moments of the mRNA distribution as we continue to
examine the discriminatory power of these different models.
